One Model May Not Fit All: Subgroup Detection Using Model-Based Recursive Partitioning

### 2023-10-20 _MOB_paper.Rnw
text MOB intro, fig 1 and 2 in one

\title{One Model May Not Fit All: Subgroup Detection Using MOB}
\shorttitle{shorttitle}

@@

\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups of persons showing different effects in a wide range of parametric models. It provides a versatile tool for detecting and explaining heterogeneity of intervention effects. In this tutorial paper, we provide an introduction to the general MOB framework. In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item response theory models. In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models. We apply GLMM trees to longitudinal data from a study on the effects of Head Start, to identify subgroups of families where children show comparatively larger or smaller gains in performance. In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, i.e.~differential item functioning (DIF). %DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness. % Test fairness is more related to use of test scores in practice than group-level comparisons I would think?
We show how a recently developed stopping criterion \citep{HenDebStr:2023:EPM} can be used to guide subgroup detection based on DIF effect sizes.\\}

@@

The framework of MOB is very flexible and it can be applied to a wide range of statistical models, such as linear and generalized linear regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS,WiedyFrick21}. In this article, we will highlight two specific MOB methods that we consider particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and MOB for Rasch measurement and Item Response Theory models \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}. Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures, for example when children are tested at several time points (so that time points are nested in children) and/or when children from different classes from different schools (so that children are nested in classes, which are in turn nested in schools) participate in a study. In application example 1 \ref{sec:TutorialMixed} we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure. In this example, the focus is on detecting subgroups with different parameters of a \textit{regression} model. We thus assume that the psychological test score(s) of interest \todo{(Mirka: I like the term instrument to distinguish it from a statistical test) (Marjolein: I hope 'psychological test score' clears up possible confusion)} have already been validated.

For application example 2\todo{add link to section? not sure how to do}, we go back in the research process to the point where a new test has been administered to a validation sample. Here, we focus on detecting subgroups with different parameters of a \textit{measurement} model. Validity assessment requires that we make sure that test results are comparable, for example between children of different genders. If certain items show different measurement parameters between groups, this may put certain groups at a relative disadvantage, and these items are said to violate measurement invariance or to exhibit differential item functioning (DIF). In order to test for DIF in the framework of Item Response Theory or Rasch modelling \citep{AnthyDiPe16,DebStrZei:2022:CRC,Mall97}, the item parameters are compared between groups of persons. This can be done in a way that allows to detect DIF, while accounting for possible true group differences in ability. Traditional approaches for testing DIF require the groups to be pre-specified in order to test for DIF. In application example 2 we show how MOB flexibly allows to detect groups with different item parameters in a data-driven way.

In the following section, we first give a short introduction into the algorithm and statistical concepts behind MOB. Readers interested in learning more about its predecessor method, classification and regression trees, are referred to the introduction by \citet{StrMalTut:2009:PM}. An alternative framework for detecting groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}. Mixture modelling aims at identifying latent classes of persons with different properties. It can also be combined with covariates. A comparison of both approaches is given by \citet{FriStrZei:2014}. \todo{Last 5 sentences seem distracting here. Omit or move to discussion?}

@@

\begin{figure}
\caption{Univariate distribution of the response variable, with global mean indicated by triangle.}
\begin{subfigure}{.5\textwidth}
<<eval = TRUE, echo = FALSE, fig=TRUE, width=5, height=5>>=
library("colorspace")
set.seed(3)
x1 <- runif(250, min = 5, max = 11)
x2 <- runif(250, min = 2, max = 8)
x3 <- runif(250, min = 4, max = 10)
error <- rnorm(250, sd = 1.5)
y <- 7 + ifelse(x1 > 8 & x2 > 5, yes = 2.5, no = -2.5) + error
toy_data <- round(data.frame(x1, x2, x3, y), digits = 2L)
boxplot(toy_data$y, cex = .7)
points(x = 1, y = mean(toy_data$y), col = rainbow_hcl(1), pch = 17)
@
\end{subfigure}
\label{fig:global}
\end{figure}

@@

We illustrate the idea using a very simple toy dataset, comprising \Sexpr{nrow(toy_data)} observations, and four variables: A continuous response variable $y$, and three possible partitioning (or predictor) variables $x_1$, $x_2$ and $x_3$. To keep the example simple, we apply MOB to a very simply global model, which comprises only an intercept. Figure~\ref{fig:global} shows the distribution of $y$ in a boxplot, with the global intercept depicted as a triangle. Obviously, the global intercept does not describe all observations equally well, there is quite some unexplained variation around it.

@@

We applied MOB to the observations in Figure~\ref{fig:global}, specifying $x_1$, $x_2$ and $x_3$ as the additional covariates. The resulting tree is shown in Figure~\ref{fig:tree}. In the root node, which contains all observations, structural change tests were performed for each of the three covariates (step 2). The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected for splitting (step 3). The $p$ value resulting from the structural change test for $x_1$ is depicted in this root node. Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible between the groups. In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with instability in this subgroup. In the right daughter node, there was significant instability associated with $x_2$, a categorical covariate. Again, the cutpoint was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible. The subgroup-specific model parameters are presented in end nodes at the bottom of the tree.

\begin{figure}
\caption{Decision tree with group-specific parameters in the terminal nodes.}
\begin{subfigure}{.5\textwidth}
<<eval = TRUE, echo = FALSE, fig=TRUE, width=5, height=5>>=
library("partykit")
toy_data$x2 <- factor(ifelse(x2 > 5, yes = ifelse(x2 > 6.5, "A", "C"), no = "B"))
tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
plot(tree, gp = gpar(cex = .7))
@
~
<<eval = TRUE, echo = FALSE, fig=TRUE, width=5, height=5>>=
plot(tree, gp = gpar(cex = .7), type = "simple")
@
\end{subfigure}
\label{fig:tree}
\end{figure}

@@

There are two main differences between MOB and and traditional classification and regression tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993}. The first difference relates to the separation of variable and cutpoint selection. The traditional algorithms performed variable and cutpoint selection in one step, which leads to an undesirable behavior called variable selection bias. That is, the traditional algorithms prefer variables offering more possible cutpoints in the selection process -- regardless of their true information content. An algorithm that still has this problem is, for example, the \texttt{rpart} algorithm in R, based on the original CART algorithm by \citet{Breetal:1984}. More recent algorithms for classification and regression trees have solved this problem and offer unbiased variable selection, such as \texttt{QUEST} \citep{LohShi:1997}, which is available in SPSS, and \texttt{ctree} \citep{Hotetal:2006}, which is available in \texttt{R} in packages \texttt{party} and \texttt{partykit} \citep{partykit:pkg}. The latter forms the basis for all MOB approaches presented in this paper. For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006,StrMalTut:2009:PM,StroyKopf15}.

The second difference relates to the way MOB stops splitting: It uses a criterion of statistical significance. Once there are no more covariates that show a significant structural change in any node, splitting is halted. In this way, the MOB algorithm selects only those variables that are relevant for distinguishing the groups, i.e., it performs automatic variable selection. Moreover, the trees will not grow as large as possible, but will stop when no more significant structural change is detected. While the traditional algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993} grew very large trees and then cut them back (i.e., \textit{pruning}), the MOB framework employs significance tests, but it can also (partly) use effect size measures, as we will see in application example 2 \todo{include link}. This allows for stopping tree growing as soon as no significant differences can be detected anymore. Other stopping criteria are based on the number of persons in the end nodes. These criteria ensure that the sample sizes in the end nodes are large enough to estimate the statistical model in each end node.\todo{we should discuss this in the application examples; there are some defaults but for models with many parameters it might make sense to increase them}

An important characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory manner, but is learned from the data in an exploratory manner. This is a key feature of the MOB approach that makes it very flexible and sets it apart from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered. While there are phases in psychological and educational research where it is very important to specify hypotheses a-priori and test them in a confirmatory manner, in early stages of research exploratory methods are an important addition to the statistical toolbox for researchers. Still, an important challenge for the researcher remains: To specify the parametric model of interest, to specify the set of covariates or possible partitioning variables and to choose the settings of the MOB algorithm. The current paper aims to provide some guidance.

@@

Validity assessment of scores on psychological or educational tests requires researchers to assess whether the same construct is measured in the same way for different groups. In the framework of IRT and Rasch measurement, test items are typically investigated with respect to item misfit, multidimensionality and other violations of the measurement model \citep[cf., e.g.,][for an introduction]{DebStrZei:2022:CRC}. \todo{Omit last sentence or move to discussion? I think better to focus on DIF only here.} Another assumption that is particularly crucial for the comparability of test scores between groups is measurement invariance. Items that violate measurement invariance by showing different measurement properties for different groups of participants display DIF.

MOB can be used to detect DIF by means of searching for covariates that display structural change in the item parameters of the measurement model\todo{possibly confusing: covariates are supplied by user, not searched for. Suggestion: by means of testing whether the parameters of the measurement model exhibit significant instability with respect to (sets of) covariates?}. For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15} \todo{should we cite the package too or only the paper? (Marjolein: I think software should only be mentioned in the applications.)} The usage of this function will be illustrated in application example 2. We will see that, just like for the mixed-effects model trees\todo{use same names for methods everywhere}, we need to specify which variables are part of the parametric model. In the case of the Rasch model, this will be the test items. Moreover, we need to specify, which variables are made available to the MOB algorithm for selecting relevant splitting variables and cutpoints.

@@

As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice. Together with a purification step \citep[see][for details]{HenDebStr:2023:EPM} \todo{Caro can add more text on achoring, incl. word anchoring, here to prepare example 2 - only shortly mention here and explain in more detail later in the example (will become more clear in example)}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically. This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items. 
\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
For example, items that show DIF between different language groups can often be improved by means of making sure that in all translations the meaning is as similar as possible, that the words employed in the translations for the different languages are equally frequently used, etc. In other situations, sources of DIF might be harder to eliminate, so that often DIF items are excluded from a test. Either way, the measurement model needs to be refitted and its assumptions checked again after the final set of items has been decided upon and, in the case of modified items, administered again.

@@

\todo{code for figure not displayed on purpose? why only those particular x axis ticks?}

@@

>>>>

\title{One Model May Not Fit All: Subgroup Detection Using Model-Based Recursive Partitioning}
\shorttitle{shorttitle}

@@

\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups of persons showing different effects in a wide range of parametric models. It provides a versatile tool for detecting and explaining heterogeneity of intervention effects. In this tutorial paper, we provide an introduction to the general MOB framework. In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item response theory (IRT) models. In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models. We apply GLMM trees to longitudinal data from a study on the effects of Head Start, to identify subgroups of families where children show comparatively larger or smaller gains in performance. In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, i.e.~differential item functioning (DIF). %DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness. % Test fairness is more related to use of test scores in practice than group-level comparisons I would think?
We show how a recently developed stopping criterion \citep{HenDebStr:2023:EPM} can be used to guide subgroup detection based on DIF effect sizes.\\}

@@

The framework of MOB is very flexible and it can be applied to a wide range of statistical models, such as linear and generalized linear regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS,WiedyFrick21}. In this article, we will highlight two specific MOB methods that we consider particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and MOB for Rasch measurement and Item Response Theory models \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}. Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures, for example when children are tested at several time points (so that time points are nested in children) and/or when children from different classes from different schools (so that children are nested in classes, which are in turn nested in schools) participate in a study. In application example 1 \ref{sec:TutorialMixed} we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure. In this example, the focus is on detecting subgroups with different parameters of a \textit{regression} model. We thus assume that the psychological test score(s) of interest have already been validated. 
%\todo{(Caro: I have kept psychological test score, because instrument to me sounds more clinical)(Mirka: I like the term instrument to distinguish it from a statistical test) (Marjolein: I hope 'psychological test score' clears up possible confusion)}

For application example 2, we go back in the research process to the point where a new psychological test has been administered to a validation sample. Here, we focus on detecting subgroups with different parameters of a \textit{measurement} model. Validity assessment requires that we make sure that test results are comparable, for example between children of different genders. If certain items show different measurement parameters between groups, this may put certain groups at a relative disadvantage, and these items are said to violate measurement invariance or to exhibit differential item functioning (DIF). In order to test for DIF in the framework of Item Response Theory or Rasch modelling \citep{AnthyDiPe16,DebStrZei:2022:CRC,Mall97}, the item parameters are compared between groups of persons. This can be done in a way that allows to detect DIF, while accounting for possible true group differences in ability. Traditional approaches for testing DIF require the groups to be pre-specified in order to test for DIF. In application example 2 we show how MOB flexibly allows to detect groups with different item parameters in a data-driven way.

In the following section, we first give a short introduction into the algorithm and statistical concepts behind MOB. Readers interested in learning more about its predecessor method, classification and regression trees, are referred to the introduction by \citet{StrMalTut:2009:PM}.

@@

@@

<<eval = TRUE, echo = FALSE>>=
set.seed(3)
x1 <- runif(250, min = 5, max = 11)
x2 <- runif(250, min = 2, max = 8)
x3 <- runif(250, min = 4, max = 10)
error <- rnorm(250, sd = 1.5)
y <- 7 + ifelse(x1 > 8 & x2 > 5, yes = 2.5, no = -2.5) + error
toy_data <- round(data.frame(x1, x2, x3, y), digits = 2L)
toy_data$x2 <- factor(ifelse(x2 > 5, yes = ifelse(x2 > 6.5, "A", "C"), no = "B"))
@

We illustrate the idea using a very simple, simulated toy dataset, comprising \Sexpr{nrow(toy_data)} observations and four variables: A continuous response variable $y$, and three covariates, $x_1$, $x_2$ and $x_3$, as possible partitioning variables. 
%note from Caro: I woul call them partitioning variables or covariates, but not predictors, to save the term predictors for those x-variables that are part of the parametric model
To keep the example simple, we apply MOB to a very simple global model, which comprises only an intercept. It would also be possible to use, e.g., a linear regression model or a logistic regression model as the global model. Figure~\ref{fig:toy}, left, shows the distribution of $y$ in a boxplot, with the global intercept indicated as a triangle. Obviously, the global intercept does not describe all observations equally well, there is quite some unexplained variation around it.
%note from Caro: I found Fig 1 containing only the boxplot irritating and have moved it into one panel with the tree

@@

We applied MOB to the observations in Figure~\ref{fig:toy}, left, specifying $x_1$, $x_2$ and $x_3$ as potential splitting variables. The resulting tree is shown in Figure~\ref{fig:toy}, right. In the root node, which contains all observations (step 1), structural change tests were performed for each of the three covariates (step 2). The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected first for splitting (step 3). The $p$ value resulting from the structural change test for the first split in $x_1$ is depicted in the root node. Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit the strongest parameter differences. In the right daughter node, additional significant instability was detected with respect to $x_2$, a categorical covariate. Again, the cutpoint in $x_2$ was selected so that the two resulting subgroups exhibit strongest parameter differences. In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with any further instability in this subgroup (step 4). The same held for nodes 4 and 5, so that the third covariate, $x_3$, was never selected for splitting. The subgroup-specific distributions of the response variable are presented in end nodes at the bottom of the tree.
The fact that Figure~\ref{fig:toy}, right, displays more than one end node confirms that one global model for all observations cannot appropriately capture the pattern in the data. 
Figure~\ref{fig:toy}, right, also shows that, out of the three covariates that were presented to the algorithm, only two were actually selected for splitting. This automatic variable selection is an important characteristic of classification and regression tree and MOB algorithms.

\begin{figure}
\caption{Left: Univariate distribution of the response variable, with global mean indicated by triangle. Right: Tree with group-specific distributions of the response variable in the terminal nodes.}
\begin{subfigure}[][][t]{.4\textwidth}
<<eval = TRUE, echo = FALSE, fig=TRUE, width=4.5, height=7>>=
library("colorspace")
boxplot(toy_data$y, cex = .7)
points(x = 1, y = mean(toy_data$y), col = rainbow_hcl(1), pch = 17, ylim = c(1,14))
@
\end{subfigure}
\begin{subfigure}[][][b]{.6\textwidth}
<<eval = TRUE, echo = FALSE, fig=TRUE, width=6, height=5>>=
library("partykit")
tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
plot(tree, gp = gpar(cex = .7), ylim = c(1,14))
@
<<eval = FALSE, echo = FALSE, fig=TRUE, width=5, height=5>>=
plot(tree, gp = gpar(cex = .7), type = "simple")
@
\end{subfigure}
\label{fig:toy}
\end{figure}

@@

%note from Caro: I did not like the formulation "There are two main differences between MOB and and traditional classification and regression tree algorithms" because readers might misinterpret it and think that only MOB, not trees, can be unbiased and use significance tests, while the distinction is really between traditional and modern algorithms for both trees and mob, changed into: characteristics
There are three further important characteristics of tree and MOB algorithms that we would like to mention here: The first characteristic is the type of variable and cutpoint selection an algorithm employs. Traditional classification and regression tree algorithms, like those of \citet{Breetal:1984} and \citet{Qui:1993}, performed variable and cutpoint selection in one step, which leads to an undesirable behavior called variable selection bias. That is, the traditional algorithms prefer variables offering more possible cutpoints in the selection process -- regardless of their true information content. An algorithm that still has this problem is, for example, the \texttt{rpart} algorithm in R, based on the original CART algorithm by \citet{Breetal:1984}. More modern algorithms for classification and regression trees have solved this problem and offer unbiased variable selection, such as \texttt{QUEST} \citep{LohShi:1997}, which is available in SPSS, and \texttt{ctree} \citep{Hotetal:2006}, which is available in \texttt{R} in packages \texttt{party} and \texttt{partykit} \citep{partykit:pkg}. The latter forms the basis for all MOB approaches presented in this paper. For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006,StrMalTut:2009:PM,StroyKopf15}.

The second characteristic relates to the way a tree or MOB algorithm stops splitting: Modern algorithms for classification and regression trees and MOB use a criterion of statistical significance to stop splitting. Once there are no more covariates that show a significant structural change in any node, splitting is halted. In this way, the  algorithm selects only those partitioning variables that are relevant for distinguishing the groups (i.e., it performs automatic variable selection, as illustrated in Figure~\ref{fig:toy}, right). Moreover, the trees will not grow as large as possible, but will stop when no more significant structural change is detected. While traditional tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993} grew very large trees and then cut them back (so called \textit{pruning}), modern tree and MOB algorithms employ significance tests as stopping criteria (and some can also use effect size measures, as we will see in application example 2). This allows for stopping tree growing as soon as no significant differences can be detected anymore. Other stopping criteria are based on the number of persons in the end nodes. These criteria ensure that the sample sizes in the end nodes are large enough to estimate the statistical model in each end node.\todo{we should discuss this in the application examples; there are some defaults but for models with many parameters it might make sense to increase them}

The third important characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory manner, but is learned from the data in an exploratory manner. This is a key feature of the MOB approach that makes it very flexible and sets it apart from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered. While there are phases in psychological and educational research where it is very important to specify hypotheses a-priori and test them in a confirmatory manner, in early stages of research exploratory methods are an important addition to the statistical toolbox for researchers. Still, an important challenge for the researcher remains: To specify the parametric model of interest, to specify the set of possible partitioning variables and to choose the settings of the MOB algorithm. The current paper aims to provide some guidance. 
%note from Caro: it said include link in several places when referring to example 2, but I don't think this is necessary, because currently sections are not numbered anyway and the title example 2 is clear

@@

Validity assessment of scores on psychological or educational tests requires researchers to assess whether the same construct is measured in the same way for different groups. In the framework of IRT and Rasch measurement, test items are typically investigated with respect to item misfit, multidimensionality and other violations of the measurement model \citep[cf., e.g.,][for an introduction]{DebStrZei:2022:CRC}. \todo{Omit last sentence or move to discussion? I think better to focus on DIF only here. Response Caro: would rather keep it here, but have increased importance of measurement invariance in next sentence.} A particularly crucial assumption for the comparability of test scores between groups is measurement invariance. Items that violate measurement invariance by showing different measurement properties for different groups of participants display DIF.

MOB can be used to detect DIF by means of testing whether the item parameters of the measurement model exhibit significant instability with respect to (combinations of) covariates. For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15}. The usage of this function will be illustrated in application example 2. We will see that, just like for  GLMM trees, we need to specify which variables are part of the parametric model. In the case of the Rasch model, this will be the test items. Moreover, we need to specify, which covariates are made available to the MOB algorithm for selecting relevant splitting variables and cutpoints.

@@

As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice. Together with a purification step and anchoring \citep[see][and application example 2 for details]{HenDebStr:2023:EPM}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically. This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items. 
%\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
For example, items that show DIF between different language groups can often be improved by means of making sure that in all translations the meaning is as similar as possible, that the words employed in the translations for the different languages are equally frequently used, etc. In other situations, sources of DIF might be harder to eliminate, so that often DIF items are excluded from a test. Either way, the measurement model needs to be refitted and its assumptions checked again after the final set of items has been decided upon and, in the case of modified items, administered again.

@@

@@

\todo{next paragraph moved to here, where exactly?}
An alternative framework for detecting groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}. Mixture modelling aims at identifying latent classes of persons with different properties. It can also be combined with covariates. A comparison of both approaches is given by \citet{FriStrZei:2014}.

### 2023-10-20 _MOB_paper.tex
text MOB intro, fig 1 and 2 in one

\begin{figure}
\caption{Univariate distribution of the response variable, with global mean indicated by triangle.}
\begin{subfigure}{.5\textwidth}
\includegraphics{_MOB_paper-002}
\end{subfigure}
\label{fig:global}
\end{figure}

@@

We illustrate the idea using a very simple toy dataset, comprising 250 observations, and four variables: A continuous response variable $y$, and three possible partitioning (or predictor) variables $x_1$, $x_2$ and $x_3$. To keep the example simple, we apply MOB to a very simply global model, which comprises only an intercept. Figure~\ref{fig:global} shows the distribution of $y$ in a boxplot, with the global intercept depicted as a triangle. Obviously, the global intercept does not describe all observations equally well, there is quite some unexplained variation around it.

@@

We applied MOB to the observations in Figure~\ref{fig:global}, specifying $x_1$, $x_2$ and $x_3$ as the additional covariates. The resulting tree is shown in Figure~\ref{fig:tree}. In the root node, which contains all observations, structural change tests were performed for each of the three covariates (step 2). The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected for splitting (step 3). The $p$ value resulting from the structural change test for $x_1$ is depicted in this root node. Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible between the groups. In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with instability in this subgroup. In the right daughter node, there was significant instability associated with $x_2$, a categorical covariate. Again, the cutpoint was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible. The subgroup-specific model parameters are presented in end nodes at the bottom of the tree.

\begin{figure}
\caption{Decision tree with group-specific parameters in the terminal nodes.}
\begin{subfigure}{.5\textwidth}
\includegraphics{_MOB_paper-003}
~
\includegraphics{_MOB_paper-004}
\end{subfigure}
\label{fig:tree}
\end{figure}

@@

\begin{figure}%
\caption{}
\begin{subfigure}{.7\textwidth}
\includegraphics{_MOB_paper-008}
\end{subfigure}
\label{fig:global_lmm}
\end{figure}%

\todo{code for figure not displayed on purpose? why only those particular x axis ticks?}

@@

\begin{figure}%
\caption{}
\includegraphics{_MOB_paper-011}

@@

\includegraphics{_MOB_paper-012}
\label{fig:lmm_tree}
\end{figure}%

@@

\begin{Schunk}
\begin{Sinput}
 plot(Raschtree_culture)
\end{Sinput}
\end{Schunk}
\includegraphics{_MOB_paper-020}

@@

\begin{Schunk}
\begin{Sinput}
 plot(Raschtree_MH_culture)
\end{Sinput}
\end{Schunk}
\includegraphics{_MOB_paper-025}

@@

\begin{Schunk}
\begin{Sinput}
 plot(Raschtree_MH_culture, color_by_node = 1)
\end{Sinput}
\end{Schunk}
\includegraphics{_MOB_paper-026}

@@

\begin{Schunk}
\begin{Sinput}
 plot(Raschtree_MH_culture, color_by_node = 2)
\end{Sinput}
\end{Schunk}
\includegraphics{_MOB_paper-027}

@@

\begin{Schunk}
\begin{Sinput}
 plot(Raschtree_MH_culture, color_by_node = 4)
\end{Sinput}
\end{Schunk}
\includegraphics{_MOB_paper-028}

@@

\begin{figure}
\caption{Association between Age and PPVT scores before (left) and after (right) transformation.}
\includegraphics{_MOB_paper-030}
\end{figure}

>>>>

@@

We illustrate the idea using a very simple, simulated toy dataset, comprising 250 observations and four variables: A continuous response variable $y$, and three covariates, $x_1$, $x_2$ and $x_3$, as possible partitioning variables. 
%note from Caro: I woul call them partitioning variables or covariates, but not predictors, to save the term predictors for those x-variables that are part of the parametric model
To keep the example simple, we apply MOB to a very simple global model, which comprises only an intercept. It would also be possible to use, e.g., a linear regression model or a logistic regression model as the global model. Figure~\ref{fig:toy}, left, shows the distribution of $y$ in a boxplot, with the global intercept indicated as a triangle. Obviously, the global intercept does not describe all observations equally well, there is quite some unexplained variation around it.
%note from Caro: I found Fig 1 containing only the boxplot irritating and have moved it into one panel with the tree

@@

We applied MOB to the observations in Figure~\ref{fig:toy}, left, specifying $x_1$, $x_2$ and $x_3$ as potential splitting variables. The resulting tree is shown in Figure~\ref{fig:toy}, right. In the root node, which contains all observations (step 1), structural change tests were performed for each of the three covariates (step 2). The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected first for splitting (step 3). The $p$ value resulting from the structural change test for the first split in $x_1$ is depicted in the root node. Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit the strongest parameter differences. In the right daughter node, additional significant instability was detected with respect to $x_2$, a categorical covariate. Again, the cutpoint in $x_2$ was selected so that the two resulting subgroups exhibit strongest parameter differences. In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with any further instability in this subgroup (step 4). The same held for nodes 4 and 5, so that the third covariate, $x_3$, was never selected for splitting. The subgroup-specific distributions of the response variable are presented in end nodes at the bottom of the tree.
The fact that Figure~\ref{fig:toy}, right, displays more than one end node confirms that one global model for all observations cannot appropriately capture the pattern in the data. 
Figure~\ref{fig:toy}, right, also shows that, out of the three covariates that were presented to the algorithm, only two were actually selected for splitting. This automatic variable selection is an important characteristic of classification and regression tree and MOB algorithms.

\begin{figure}
\caption{Left: Univariate distribution of the response variable, with global mean indicated by triangle. Right: Tree with group-specific distributions of the response variable in the terminal nodes.}
\begin{subfigure}[][][t]{.4\textwidth}
\includegraphics{_MOB_paper-003}
\end{subfigure}
\begin{subfigure}[][][b]{.6\textwidth}
\includegraphics{_MOB_paper-004}
\end{subfigure}
\label{fig:toy}
\end{figure}

@@

\begin{figure}%
\caption{}
\begin{subfigure}{.7\textwidth}
\includegraphics{_MOB_paper-009}
\end{subfigure}
\label{fig:global_lmm}
\end{figure}%

@@

\begin{figure}%
\caption{}
\includegraphics{_MOB_paper-012}

@@

\includegraphics{_MOB_paper-013}
\label{fig:lmm_tree}
\end{figure}%

@@

\begin{Schunk}
\begin{Sinput}
 plot(Raschtree_culture)
\end{Sinput}
\end{Schunk}
\includegraphics{_MOB_paper-021}

@@

\begin{Schunk}
\begin{Sinput}
 plot(Raschtree_MH_culture)
\end{Sinput}
\end{Schunk}
\includegraphics{_MOB_paper-026}

@@

\begin{Schunk}
\begin{Sinput}
 plot(Raschtree_MH_culture, color_by_node = 1)
\end{Sinput}
\end{Schunk}
\includegraphics{_MOB_paper-027}

@@

\begin{Schunk}
\begin{Sinput}
 plot(Raschtree_MH_culture, color_by_node = 2)
\end{Sinput}
\end{Schunk}
\includegraphics{_MOB_paper-028}

@@

\begin{Schunk}
\begin{Sinput}
 plot(Raschtree_MH_culture, color_by_node = 4)
\end{Sinput}
\end{Schunk}
\includegraphics{_MOB_paper-029}

@@

\begin{figure}
\caption{Association between Age and PPVT scores before (left) and after (right) transformation.}
\includegraphics{_MOB_paper-031}
\end{figure}
